\documentclass{article}

\input{includes/commands.tex}
\input{includes/theoremstyle.tex}
\usepackage{tikz}
\usepackage[colon, sort&compress]{natbib}
\title{Preliminaries}

\date{\today}

\begin{document}
% \author{\aut}
\maketitle

\section{Preliminaries}
In this course, we use the terms network and graph interchangeably. A graph/network is a pair $G=(V, E)$, where $E \subseteq V \times V$. An edge is thus a pair $(u,v) \in E$. 

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[scale=1.2, every node/.style={circle, draw, minimum size=0.5cm}]
  \node (A) at (0,0) {1};
  \node (B) at (2,0) {3};
  \node (C) at (1,1) {2};
  \node (D) at (1,-1) {4};
  \node (E) at (1,-2) {5};
  
  % \draw (A) -- (B) node[midway, below] {Edge 1};
  % \draw (B) -- (C) node[midway, right] {Edge 2};
  % \draw (A) -- (C) node[midway, above, sloped] {Edge 3};
  \draw (A) -- (B) ;
  \draw (B) -- (C) ;
  \draw (A) -- (C) ;
  \draw (B) -- (D) ;
  \draw (A) -- (D) ;
  \draw (E) -- (D) ;
\end{tikzpicture}
    \caption{A graph with $|V|=5$}
    \label{fig: example graph}
\end{figure}

\begin{definition}
    A \textbf{path} in a graph is an alternating sequence of vertices and edges given by 
    \[
    \mathcal{P} = (v_0, e_1, v_1, e_2, \ldots, e_n,v_n)
    \]
    If there is a path between every pair of vertices, the graph is called \textbf{connected}.   
\end{definition}  

The length of a path is the \textbf{number of edges} in the path.

\subsection{Representations} 

In this course, we use the adjacency matrix to represent graphs. The following definiton (taken from the lectures) is for a simple graph, where there is at most one edge between any pair of vertices.  

\begin{definition}
    The \textbf{adjacency matrix} of a graph $G=(V,E)$ is a $n\times n$ matrix $A$ with each entry $(A)_{uv}$ for each pair of vertices $(u,v)\in E$ defined as:
    \begin{singlespace}
        \begin{equation*}
         (A)_{uv} = \begin{cases}
            1 & \mathrm{if} \ u\sim v \\
            0 & \mathrm{otherwise}
        \end{cases}
    \end{equation*}
    \end{singlespace}
    where we denote $n=|V|$.
\end{definition}  

In the example above, the adjacency matrix is given by:

\begin{equation*}
        A=\begin{pmatrix}
         0&  1&  1&  1& 0\\
         1&  0&  1&  0& 0\\
         1&  1&  0&  1& 0\\
         1&  0&  1&  0& 1\\
         0&  0&  0&  1& 0\\
    \end{pmatrix}
    % \caption{Adjacency matrix for $\mathcal{G}$}
    \label{tab:adj}
\end{equation*}


Some commonly discussed (deterministic) graphs include
\begin{itemize}
    \item \textbf{The complete graph} has $\binom{n}{2}$ edges, with an adjacency matrix with all entries being $1$ except for the diagonal entries. 
    \item \textbf{The} $k$\textbf{-partite graph} see \citep{DiestelReinhard2017GT}
    \item \textbf{The star graph} is a $K_{1,k}$ partite graph, with $v_1$ being the central vertex and $v_2, \ldots, v_{k-1}$ being the vertices attached to it. The 2-star is a triangle missing an edge. 
\end{itemize}


\begin{figure}[H]
    \centering
    \begin{tikzpicture}
% Define the left set of vertices
\foreach \x in {1,2,3,4}
    \node[draw, circle] (\x) at (0, -\x) {\x};
% Define the right set of vertices
\foreach \y in {5,6,7}
    \node[draw, circle] (\y) at (2, -\y+4) {\y};
% Add edges between the sets
\foreach \x in {1,2,3,4}
    \foreach \y in {5,6,7}
        \draw (\x) -- (\y);
\end{tikzpicture}
    \caption{A bipartite graph $K_{4,3}$}
    \label{fig: example bipartite}
\end{figure}


They admit adjacency matrices with special structures.

\subsection{Summary Statistics}
When dealing with large graphs, it would be convenient to have simple statistics that reveal information about the underlying structure.  

\begin{itemize}
    \item  The \textbf{density} of a network is given by $\frac{2e}{n(n-1)}$
    \item The \textbf{average degree} of a network is given by $\bar{d}=\frac{1}{n}\sum_{v\in V} \sum_{u\in V, u\neq v} a_{uv}$, where $a_{uv}$ is the entry in the adjacency matrix
    \item The \textbf{degree sequence} of a given network on $V$ is the set of degrees $\{d(v): v\in V\}$
    \item The \textbf{degree distribution} is a vector with each element $d_k=\frac{1}{n} |D_k|$, where $D_k$ is the set of vertices with degree $k$
\end{itemize}
\begin{remark}
    In Sheet 1, we will see that given a sequence of non-negative integers, there is an algorithm (\textbf{Havel-Hakimi}) to determine if this sequence is a degree sequence. (if so we call it graphical)
\end{remark}

\begin{definition}
    The \textbf{global clustering coefficient} (often abbreviated as just the clustering coefficient) is defined as
    \begin{equation*}
        C = \frac{\sum_{u,v,w \in V} a_{uv}a_{vw}a_{wu}}{\sum_{v \in V} d(v)(d(v)-1)}
    \end{equation*}
    where the numerator is $6\times$the number of triangles and the denominator is the number of paths of length $2$:
    \[
        C = \frac{6\times \text{number of triangles}}{\text{number of 2-paths}}
    \]
    Also, note that the number $6$ arises from the fact that we have $3!=6$ ways to choose an ordered triplet.
\end{definition}

 Note that if we divide both by $2$, then the denominator is the number of $2-$stars in the graph.

A similar statistic is the local clustering coefficient for a particular vertex $v$.

\begin{definition}
    The \textbf{local clustering coefficient} of $v$ is given by:
    \begin{equation*}
        C(v) =  \frac{\sum_{u,w \in V} a_{uv}a_{vw}a_{wu}}{d(v)(d(v)-1)} 
        % = \frac{\frac{1}{2}\sum_{u,w \in N(v), u\neq w} a_{uv}a_{wv}}{\binom{d(v)}{2}}
    \end{equation*}
    where the numerator is $2\times$the number of triangles and the denominator is the number of paths of length $2$ containing $v$:
    \[
        C(v) = \frac{2\times \text{number of triangles containing } v}{\text{number of 2-paths containing }v}
    \]
\end{definition}
Note the denominator has an alternative form $\sum_{u, w \in N(v), u\neq w} a_{uv}a_{wv}$, where $N(v)$ denotes the neighbourhood of $v$; to see this, note we first sum over all pairs $(u,w)$ and then subtract the cases where they are equal, which gives $d(v)^2-d(v)$, as desired.

Thus, the \textbf{average local clustering coefficient} is 
\[
    \overline{C}=\frac{1}{|V|}\sum_{v \in V} C(v)
\]

Aside from statistics that arise from degrees of vertices, the global structure is also captured by paths in graphs. 

\begin{definition}
    The \textbf{average shortest path} is given by:
    \[
        \ell = \frac{1}{|V|(|V|-1)} \sum_{u\neq v, u,v\in V} \ell (u,v)
    \]
    where $\ell(u,v)$ is the length of the shortest path between $u$ and $v$.
\end{definition}

\begin{remark}
    We can compute $\ell(u,v)$ from the smallest power of adjacency matrix $A$ such that the $(u,v)^{th}$ entry of $A^p$ is non-zero. (This can be shown by induction)
\end{remark}

\begin{definition}
    The \textbf{motif} of a graph is a small, over-represented subgraph, where over-representation is judged using a probabilistic model for the network.
\end{definition}

\newpage
\bibliographystyle{apalike}
\bibliography{bibliography1.bib}

\end{document}